Clustering Constraints on the Relative Sizes of Central and Satellite Galaxies

### 2017-11-06 paper/galsize_paper.tex
Revisions to section 5.2

A straightforward resolution to this puzzle is shown in Figure \ref{fig:censatsizehist}, which compares the $\rhalf$ distributions of central, satellite, and splashback galaxies with the same halo mass $\mhalo\equiv\mpeak\approx10^{12}\msun.$ A ``splashback central"  is defined as a present-day central that used to be a satellite, i.e., its main progenitor halo passed inside the virial radius of a larger halo at some point in its past history. On the other hand, we define a ``true central" as a galaxy that has never been a satellite.

@@

Upon first glance, the $\rhalf-$dependence of color- and morphology-selected samples shown in Figure \ref{fig:colorclustering} appear to be driven by different principles, since the trends substantially weaken and change sign. However, even very simple models based entirely on central vs. satellite size differences can give rise to this behavior. For example, it is well known that satellites are redder than centrals \citep[e.g.,][]{vdB_etal08}; accordingly, red galaxies cluster much stronger relative to blue \citep{zehavi_etal11}. Pre-selecting galaxy samples by color will thus significantly weaken the dynamic range of $F_{\rm sat}$ within each color-selected subsample. This reduction in the role of satellites could be sufficient for the $\rhalf\propto\rvir\propto\mhalo$ of {\em central} galaxies to become the dominant effect, which would result in the mild, reversed trend seen in the top panels of Figure \ref{fig:colorclustering}. The same mechanism could be at work for the case of morphology pre-selection, except that for disk-dominated systems, there would need to be a larger residual correlation with satellite designation relative to blue-selected samples \citep[This is, in fact, exactly the trend reported in][]{weinmann_etal08}.

While we consider such an explanation plausible, a conclusive determination of the principles underlying the clustering trends shown in Figure \ref{fig:colorclustering} requires proper forward modeling of morphology and color jointly with galaxy size. Such an effort is beyond the scope, but well-motivated, by the present work.

@@

Physical explanations for differences between the sizes of centrals and satellites fall into two categories, depending on whether or not the processes involved are post-infall and satellite-specific. The success of the $\rvir-$based model, together with the failure of the $\mstar$+stripping model, suggests that the relative size of central and satellite galaxies is largely driven by the {\rm pre-infall} evolution of the satellite. The clustering ratios in Figure \ref{fig:mstarmodelclustering} show that the impact of physically motivated levels of satellite mass loss likely cannot explain the $\rhalf-$dependent clustering seen in SDSS. As an additional test, we reach the same conclusion when implementing an even more extreme $\mstar-$stripping model in which stellar mass loss is linearly proportional to halo mass loss \citep[][Model 1]{watson_etal12}.

While these clustering measurements give supporting evidence that satellite mass stripping plays a sub-dominant role in shaping galaxy size, we point out that our treatment of satellite mass loss is only approximate: we have modified the size of satellites without self-consistently modifying total stellar mass. In the present work, we have opted not to develop and fine-tune a complex, semi-analytic model; instead, we have made a simple ``first-order" estimate for the level at which satellite-specific stripping impacts the two-point function. Our findings support the idea that the small sizes of satellites arise primarily from the satellite's distinctive evolution within a high density region and a steep tidal field {\em prior} to becoming a present-day satellite.

@@

As discussed in \S\ref{sec:future}, we are currently pursuing follow-up work in which we jointly model morphological type together with galaxy size. We consider forward modeling methods to be a requisite for meaningful progress on this issue, not only to properly handle the multi-dimensional nature of the problem, but also to rigorously treat systematic errors that plague inference based on galaxy group catalogs \citep[see][for a thorough discussion]{campbell_etal15}.

@@

{\color{blue} APH comment: Some of this discussion can/should be migrated to \S\ref{subsec:broader_implications}. Depending on how that section ends up looking, it may be that this entire Future Directions section can be deleted after cannibalization.}

@@

\subsection{Jointly modeling $\langle\mstar\vert\mhalo\rangle$}
\label{subsec:joint_smhm_modeling}

@@

\subsection{Jointly modeling morphology}
\label{subsec:joint_morphology_modeling}

A proper interpretation of the results clustering results shown in Figure \ref{fig:colorclustering} requires forward-modeling the joint dependence of the galaxy-halo connection upon $\mstar, \rhalf, g-r, $ and $B/T,$ which is beyond the scope of this work. Nonetheless, the broad features of these $\wproj(\rproj)$ measurements give insight into the characteristics that any model for these properties should exhibit.

First, the bottom right panel of Figure \ref{fig:colorclustering} indicates that the size of the bulge has only a tenuous connection to the properties of its parent dark matter halo.\footnote{Strictly speaking, the size of the bulge could in principle be closely connected to any number of dark matter halo properties, so long as those halo properties do not significantly impact two-point clustering.} Second, for disk-dominated galaxies, the observed $\rhalf-$dependent clustering suggests that at fixed $\mstar,$ the size of the disk is strongly correlated with whether or not the galaxy is a satellite. These findings imply that bulge prominence is intimately connected with the quenching of star-formation, but that there is wide variation in star-formation histories amongst disk-dominated galaxies, in agreement with numerous previous findings \citep[see, e.g.,][]{woo_etal15,font_etal17}.

Previous studies of large-scale structure measured by SDSS indicate that morphological type has a significant influence on galaxy clustering \citep{skibba_etal08}. Because morphology is covariant with star-formation rate (SFR) and $\rhalf,$ accounting for this dependency will require simultaneously modeling each of these variables. Historically, such a task has been beyond the purview of empirical models. However, recent progress has substantially improved the richness of the predictions that galaxy-halo models are capable of \citep{becker15,cohn17,moster_etal17}. This new class of models has accomplished this by empirically modeling star-formation rate, and then integrating SFR along each halo's history to differentially build up stellar mass. We are currently extending this new class of models to differentially build up the disk and bulge across cosmic time using computationally efficient empirical modeling methods. In addition to treating the multi-dimensional nature of the problem, this will allow us to use data at higher redshift in a self-consistent manner.

>>>>

A straightforward resolution to the first puzzle is shown in Figure \ref{fig:censatsizehist}, which compares the $\rhalf$ distributions of central, satellite, and splashback galaxies with the same halo mass $\mhalo\equiv\mpeak\approx10^{12}\msun.$ A ``splashback central"  is defined as a present-day central that used to be a satellite, i.e., its main progenitor halo passed inside the virial radius of a larger halo at some point in its past history. On the other hand, we define a ``true central" as a galaxy that has never been a satellite.

@@

Upon first glance, the $\rhalf-$dependence of color- and morphology-selected samples shown in Figure \ref{fig:colorclustering} appear to be driven by different principles, since the trends substantially weaken and change sign. However, even very simple models based entirely on central vs. satellite size differences can give rise to this behavior. For example, it is well known that satellites are redder than centrals \citep[e.g.,][]{vdB_etal08}; accordingly, red galaxies cluster much stronger relative to blue \citep{zehavi_etal11}. Pre-selecting galaxy samples by color will thus significantly weaken the dynamic range of $F_{\rm sat}$ within each color-selected subsample. This reduction in the role of satellites could be sufficient for the $\rhalf\propto\rvir\propto\mhalo$ of {\em central} galaxies to become the dominant effect, which would result in the mild, reversed trend seen in the top panels of Figure \ref{fig:colorclustering}. The same satellite-driven mechanism could be at work for the case of morphology pre-selection \citep[as in the morphology models developed in][]{skibba_etal08}, except that for disk-dominated systems, there would need to be a larger residual correlation with satellite designation relative to blue-selected samples \citep[which is, in fact, exactly the trend reported in][]{weinmann_etal08}.

While we consider such an explanation plausible, a conclusive determination of the principles underlying the clustering trends shown in Figure \ref{fig:colorclustering} requires proper forward modeling of morphology and color jointly with galaxy size. Such an effort is beyond the scope, but well-motivated, by the present work. See Section \ref{subsec:broader_implications} for further discussion.

@@

Physical explanations for differences between the sizes of centrals and satellites fall into two categories, depending on whether or not the processes involved are post-infall and satellite-specific. In the satellite mass loss model, satellites evolve exactly as centrals prior to the time of infall; this model assumes that {\em all} satellite size truncation occurs after the time of infall. The $\rvir-$based model assumes the opposite: namely that the size of {\em all} galaxies is statistically set by the $\rmpeak,$ the physical size of the virial radius at the time at which $\mhalo$ stops growing. This assumption is the natural extension of the abundance matching approximation that stellar mass largely stops growing at this time (including in-situ and ex-situ growth). The halos of typical satellites reach $\mpeak$ far outside the virial radius of their ultimate host halo \citep{behroozi_etal14}, and so in the $\rvir-$based model the size differences between satellites and centrals is largely in place prior to the time of satellite infall.

The clustering ratios in Figure \ref{fig:mstarmodelclustering} show that the impact of physically motivated levels of satellite mass loss likely cannot explain the $\rhalf-$dependent clustering seen in SDSS. As an additional test, we reach the same conclusion when implementing an even more extreme $\mstar-$stripping model in which stellar mass loss is linearly proportional to halo mass loss \citep[][Model 1]{watson_etal12}. The success of the $\rvir-$based model, together with the failure of the satellite mass loss model, suggests that the relative size of central and satellite galaxies is largely driven by the {\rm pre-infall} evolution of the satellite.

The broad features of the $\wproj(\rproj)$ measurements shown in Figure \ref{fig:colorclustering} give insight into the characteristics that should be exhibited by any model for the relationship between $B/T$ and $\rhalf.$ First, the bottom right panel of Figure \ref{fig:colorclustering} indicates that the size of the bulge has only a tenuous connection to the properties of its parent dark matter halo.\footnote{Strictly speaking, the size of the bulge could in principle be closely connected to any number of dark matter halo properties, so long as those halo properties do not significantly impact two-point clustering.} Second, for disk-dominated galaxies, the observed $\rhalf-$dependent clustering suggests that at fixed $\mstar,$ the size of the disk is strongly correlated with whether or not the galaxy is a satellite. These findings imply that bulge prominence is intimately connected with the quenching of star-formation, but that there is wide variation in star-formation histories amongst disk-dominated galaxies, in agreement with numerous previous findings \citep[see, e.g.,][]{woo_etal15,font_etal17}.

Recent progress has substantially improved capabilities for the richness of the predictions of empirical galaxy--halo models \citep{becker15,cohn17,moster_etal17}. This new class of models has accomplished this by empirically modeling star-formation rate, and then integrating SFR along each halo's history to differentially build up stellar mass. In ongoing follow-up work, we are currently extending this new class of models to differentially build up the disk and bulge across cosmic time using computationally efficient empirical modeling methods. In addition to treating the multi-dimensional nature of the problem, this will allow us to use data at higher redshift in a self-consistent manner.

% A possibly overkill caveat paragraph:
%While these clustering measurements give supporting evidence that satellite mass stripping plays a sub-dominant role in shaping galaxy size, we point out that our treatment of satellite mass loss is only approximate: we have modified the size of satellites without self-consistently modifying total stellar mass. In the present work, we have opted not to develop and fine-tune a complex, semi-analytic model; instead, we have made a simple ``first-order" estimate for the level at which satellite-specific stripping impacts the two-point function. Our findings support the idea that the small sizes of satellites arise primarily from the satellite's distinctive evolution within a high density region and a steep tidal field {\em prior} to becoming a present-day satellite.

@@

@@

@@

@@

Figure \ref{fig:censatsizeratios} shows that the satellites in $\mhalo\approx10^{12}\msun$ halos can be as large as $\lesssim40-50\%$ smaller than their central galaxy counterparts. Since the satellite fraction varies quite significantly with morphology and/or color, we have demonstrated that it is entirely plausible that most of the observed environmental trends of galaxy size can be understood simply in terms of central vs. satellite size differences. While it remains to be seen whether this explanation ends up comprising the whole story, one thing is now clear: it is simply not possible to reliably analyze environmental trends of galaxy $\rhalf$ without properly accounting for satellites. Because of the large systematics that are common in contemporary group-finding methods \citep{campbell_etal15}, we consider forward modeling methods to be a requisite for meaningful continued progress on this topic.\footnote{We note that group-finders may play an important role in future analyses of galaxy size. However, it was shown in \citet{campbell_etal15} that treating observed groups as if they were correctly identified dark matter halos can lead to entirely unreliable conclusions. The robust method to extract information from galaxy group catalogs is instead to generate a Monte Carlo realization of a forward model, and then run the same group-finding algorithm on the mock as was run on the data, enabling a proper apples-to-apples comparison.}
